State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources

### 2021-03-22 analysis/paper/paper.Rmd
made a few more revisions to address reveiewer comments and clean up some confusing language.
Clarified some responses to reviewer comments.

Re-knit to html and pdf versions

In this study, we describe a novel method to estimate aggregate and group escapements which incorporate all sources of known uncertainty and demonstrate that the estimation method is essentially unbiased, thus better informing conservation and management decision making. We also test whether observed nighttime passage and fallback/re-ascension rates are typically equal. Our method for estimating species-specific escapement past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of fish previously tagged with passive integrated transponder (PIT) tags in the adult ladder to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@Royle2008] which separates process variance (e.g., week to week variance in true escapement) from observation error variance (e.g., observation error at the window, or sampling variance at the trap). To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata. Further, total and weekly estimates are parsed into three origin groups: wild fish, hatchery fish with a clipped adipose fin, and unclipped hatchery fish. Estimates of escapement account for fish that migrate through the ladder outside of observation hours (nighttime passage) and those that ascend the ladder multiple times (re-ascension). Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for download or installation from [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM). To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates and compared STADEM results to the simulated "truth". We then applied this model to sp/sum Chinook Salmon and steelhead returns at Lower Granite Dam for spawn years 2010 - 2019.

@@

We used STADEM and three sources of data to estimate sp/sum Chinook Salmon and steelhead escapement at Lower Granite Dam from 2010-2019. Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of fish PIT tagged prior to arriving at Lower Granite Dam and detected in the adult fish ladder. Below, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; similar data could likely be obtained from other fish passage facilities.

@@

Daily counts of adult sp/sum Chinook Salmon and steelhead passing the observation window located on the Lower Granite Dam fish ladder were estimated and provided by the US Army Corps of Engineers. When summed, they provide an estimate of the number of fish ascending and passing Lower Granite Dam each season. Window counts were made for each species using direct in-person visual monitoring and video monitoring during daytime hours [@Hatch1994]. Direct visual monitoring occurred during peak run times (April 1 – October 31) for 16 hours per day (0400 – 2000 hours). Video monitoring occurred during the beginning and tail ends of the adult runs (March 1 – March 31 and November 1 – December) for 10 hours per day (0600 – 1600 hours) [@USACE2015]. During direct visual monitoring, observers recorded each adult (≥ 30cm), by species, passing the window for 50 minutes of each hour of operation. The sum of the daily 50-minute counts were then multiplied by 1.2 to account for the 10 minutes when fish were not counted. Daytime window counts were not expanded for fish that may have ascended the ladder outside of operational hours (i.e., nighttime passage) [@USACE2015]. Window counts were accessed through the Columbia Basin Research Data Access in Real Time (DART) website, [www.cbr.washington.edu/dart/query/adult_daily](www.cbr.washington.edu/dart/query/adult_daily), using their window count query. Counts were provided for each day the fish ladder was open to passage, and had already been expanded by 1.2 (to account for the counting during 50 minutes of each hour).

@@

All captured fish were anesthetized, identified to species, examined for existing marks or tags, measured for fork length and visually identified as wild or hatchery. The most widely used marking of hatchery fish is a clipped (removed) adipose fin, although coded wire tags are used in less than 10% of the hatchery releases. Some subset of hatchery fish are either intentionally or unintentionally released without a clipped adipose fin, and these are referred to as unclipped hatchery fish, or hatchery no-clips (HNC). For adults with intact adipose fins (unclipped), which includes wild and hatchery no-clip individuals, either a portion or all of the fish trapped (depending on the year) had scale and genetic tissue samples taken. Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012] and estimate the location of origin of wild fish using genetic stock identification [e.g., @Hargrove2019]. Prior to 2013, only fish determined to be wild in origin at the trap were sampled for scale and genetics. Starting in 2013, every unclipped sp/sum Chinook Salmon and steelhead trapped at Lower Granite Dam was genotyped to simplify collaborative logistics and better estimate the proportion of unclipped hatchery fish that appear phenotypically wild. @Camacho2018 provide further details on trap sample rates and valid sample selection. Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag. Final determination of wild, clipped hatchery, or unclipped hatchery origins were assigned using a post-hoc analysis of marks and tags, including parentage-based tagging [@Steele2013; @Steele2019]. Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game [@Camacho2018].

@@

The number of fish counted at the window, $Y^W_t$, is modeled as a (potentially) over-dispersed negative binomial process, with an expected value of $X_t^{day}$, the number of fish crossing the dam while the window is open. This is simply the total number of fish crossing that week, $X_t$, multiplied by the proportion of fish crossing while the window is open for counting, $\theta_t$. In the formula below, $p_t$ is the proportion of fish observed at the window and $r$ is the shape parameter. If $r$ is estimated to be small it provides evidence for over-dispersion, and as it grows very large, the negative binomial distribution behaves like a Poisson distribution.

@@

Our objective was to assess STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement for accuracy and precision, given a known simulated "truth", under different possible conditions. We developed those conditions from the combinations of two trap rate scenarios (constant and shut down for 3 weeks), two fallback/re-ascension and nighttime passage combinations (equal and offsetting, and unequal), and three window count error rates (none, low and high); resulting in twelve different scenarios (Table \@ref(tab:lgr-sim-tab)). The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at Lower Granite Dam between 2010-2015. We generated 99 simulations for each scenario, and ran STADEM on each one, resulting in a total of 1,188 model runs. Further details about simulation procedures can be found in [Appendix A](#append1).

To assess the performance of the model, we evaluated the estimates of origin-specific escapement by several measures. Accuracy was captured by the relative bias, the difference between the simulated "truth" and the STADEM estimate, divided by the simulated "truth". We estimated the precision by examining the average coefficient of variation (CV) of the estimates. We calculated the root mean squared error (RMSE) as the square root of the mean of the squared bias in the estimate, which is equivalent to the square root of the variance of the estimator plus the square of the expected bias. Finally, we evaluated whether estimates of uncertainty were appropriate by calculating coverage probabilities by determining what proportion of the model results generated an estimated 95% credible interval that contained the simulated true value. To compare the STADEM results to strictly window counts, we compared the model estimates of total unique fish and the window counts with the simulated truth.

@@

STADEM results showed similar patterns for estimates of unique total fish and unique wild fish, in that both were unbiased with similar variance in the estimator across all scenarios (Figure \@ref(fig:rel-bias-lgd)). For unique total fish, window counts were unbiased, except in scenarios when night passage and re-ascension rates were not equal (N-R). In those scenarios, they were biased by nearly 5%, the difference between night passage and re-ascension rates (Table \@ref(tab:lgr-sim-tab)). The variance of the window counts also grew as the simulated observer error grew (no error vs. Err_L vs. Err_H), whereas the variance in the STADEM estimates remained fairly constant, albeit often larger than the window count variance.

Estimates of wild escapement were unbiased, with an average relative bias of 0.2–0.3%. The CV of the estimates averaged 2.0–3.0%, with higher CV's being associated with scenarios when the trap was closed for 3 weeks. The coverage probabilities always exceeded 95% across all scenarios. The RMSE was near 500 for each scenario, representing an estimate within 2% of the true value, demonstrating the accuracy and precision of STADEM (Table \@ref(tab:summ-tab)).

@@

Weekly nighttime passage and re-ascension proportions were significantly different (2-sample test of proportions, *p* < 0.05) in approximately 50% of the weeks with sufficient PIT tag data to test. The differences in the estimates were also apparent visually (Figure \@ref(fig:night-reasc-diff-fig)). In particular, there were several weeks when the window counts were quite large, but the nighttime passage and re-ascension rates differed by as much as 10%.

@@

The results from our simulations demonstrate that STADEM provides consistent, unbiased estimates of escapement across a range of scenarios. Regardless of whether observer error at the window is negligible, low or high, whether nighttime passage and re-ascension rates are equal or not, whether the trap is shut down for several weeks in a season or not, STADEM produced estimates of unique escapement that were unbiased, precise (CVs in the 2-3% range) and contained appropriate credible intervals. Although the average bias was often positive across our simulation scenarios (Table \@ref(tab:summ-tab)), Figure \@ref(fig:rel-bias-lgd) shows that some simulations resulted in negative biases, while some were positive. We posit that conducting further simulations would bring the average bias closer to zero, but run times constrained how many simulations we were able to conduct.

As for application at Lower Granite Dam, STADEM weekly estimates were often slightly lower than those based on the number of fish caught in the adult trap because the trap estimates do not account for fallback and re-ascension (Figure \@ref(fig:jags-fit-fig)). However, the model's uncertainty is always smaller than the uncertainty from the trap estimates alone, whereas the window counts by themselves provide no estimate of uncertainty. When comparing the total estimates of each season with the window counts (Table \@ref(tab:lgr-tab)), STADEM estimates were higher when night passage rates exceeded re-ascension rates, and lower when the re-ascension rate was higher than the night passage rate. Given that there were several weeks with large window counts when the rates differed by as much as 10% (Figure \@ref(fig:night-reasc-diff-fig)), the counts for those weeks should be considered biased. ***For weeks when very few fish were caught in the trap or there was more uncertainty about the trap rate, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig).***

@@

STADEM is an improvement on window counts alone for several reasons. The first is that by accounting for night passage and re-ascension rates that are sometimes unequal, it corrects any bias in window counts. That bias ranged from -9.9% to 11% across all species and year combinations we fit STADEM to (Table \@ref(tab:lgr-tab)). Whether that amount of potential bias is of concern to management agencies is a decision for them, but we have presented a tool to correct that bias. The other justification for STADEM over window counts is that it provides an estimate of observer error for the window counts, leading to estimates of precision for unique fish which are unobtainable with window counts alone. Capturing and accounting for known sources of error is prudent to minimize management decision risk [@Harwood2003; @Ascough-Ii2008]. At the same time, STADEM's precision is still quite good, with CV's near 2-3%, well below the goal of 15% set forth by NOAA [@Crawford2011].

By incorporating data from the adult fish trap with live and video window counts in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly (for example, see Figure \@ref(fig:jags-fit-fig) when the trap was shut down for several weeks in July and August). At Lower Granite Dam, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently in five of the last ten years, often during peak run times [@USACE2010; @USACE2011; @USACE2012; @USACE2013; @Ogden2016a; @USACE2015; @USACE2016; @USACE2017; @USACE2018; @USACE2019]. Trap closures are typically associated with elevated water temperatures resulting in potential fish handling stress and/or trap malfunctions [@Ogden2016b]. Given predicted Pacific Northwest climate change scenarios [@Zhang2019] trap closures from high water temperatures may become more commonplace in the future, amplifying the need for a modeling framework that accounts for periods of missing data while still capturing estimate uncertainty. Additionally, having a framework in place that accounts for missing periods of data will allow for increased logistic flexibility if, for example, maintenance or construction is needed at the observational window or adult trap.

STADEM could be modified and run on a weekly basis or in near real-time to provide better in-season estimates for fisheries managers. Currently, the only roadblock to this at Lower Granite Dam is the identification of hatchery no-clip fish using genetic tissue samples [@Steele2013; @Steele2019] collected at the adult trap, which currently is completed post-hoc after the trapping season. The inclusion of genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement. However, if in-season management decisions do not require this correction or could accept the potential bias, origin calls at the trap could be used in-season as a first approximation to escapement. Final post-hoc estimates parsed by origin could then be finalized at season’s end. All other data included in this model (e.g., window counts and PIT observations) are otherwise provided in near real-time by DART. Provided the Lower Granite Dam adult fish trap database was updated and available in near real-time, there are minimal obstacles for adapting the STADEM framework to provide in-season estimates of escapement.

Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate escapement of sp/sum Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@IPTDSW2020; @Kinzer2020]. Estimates of species and origin-specific escapement at Lower Granite Dam, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates [e.g., @Camacho2018; @Schrader2013] that are important for fisheries management and productivity monitoring of wild populations. As an example, STADEM is being applied at Lower Granite Dam to estimate the total unique wild fish migrating past the dam. Estimates of fish passing the dam are then combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @Waterhouse2020, to estimate escapement to Snake River populations and locations throughout the basin [@Orme2019]. Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations. With sex and age data collected at the adult fish trap [@Hargrove2019], this approach provides necessary information to evaluate productivity and population viability for select Snake River sp/sum Chinook Salmon and steelhead groups [@IPTDSW2020].

With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall-run Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run and tag detection infrastructure (e.g., a PIT tag detection array or similar). Justification and infrastructure exist for applying a modified STADEM framework for fish passing Bonneville Dam, the lowest dam on the Columbia River, or Priest Rapids Dam in the upper Columbia River. Both locations currently trap a sub-sample of passing Chinook Salmon and/or steelhead for biological information and use window counts as a surrogate of true escapement. However, each has at least some similar problems to those observed at Lower Granite Dam such as unaccounted observer and sampling error, nighttime passage, and/or re-ascension. Certainly, estimating an unbiased total return to the entire Columbia River basin (i.e., Bonneville Dam) and Upper Columbia River with uncertainty would benefit managers and decision making.

@@

```{r dag-fig, fig.height = 5, fig.cap = "Directed acyclic graph showing the STADEM model framework."}
include_graphics("../figures/ModelSchematic.png")
```

>>>>

In this study, we describe a novel method to estimate aggregate and group escapements which incorporate all sources of known uncertainty and demonstrate that the estimation method is essentially unbiased, thus better informing conservation and management decision making. We also test whether observed nighttime passage and fallback/re-ascension rates are typically equal. Our method for estimating species-specific escapement past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of fish previously tagged with passive integrated transponder (PIT) tags in the adult ladder to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@Royle2008] which separates process variance (e.g., week to week variance in true escapement) from observation error variance (e.g., observation error at the window, or sampling variance at the trap). To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata. Further, total and weekly estimates are parsed into three origin groups: wild fish, hatchery fish with a clipped adipose fin, and unclipped hatchery fish. Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for download or installation from [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM). To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates and compared STADEM results to the simulated "truth". We then applied this model to sp/sum Chinook Salmon and steelhead returns at Lower Granite Dam for spawn years 2010 - 2019.

@@

Data sources for STADEM included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of fish PIT tagged prior to arriving at Lower Granite Dam and detected in the adult fish ladder. Below, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; similar data could likely be obtained from other fish passage facilities.

@@

Daily counts of adult sp/sum Chinook Salmon and steelhead passing the observation window located on the Lower Granite Dam fish ladder were compiled and provided by the US Army Corps of Engineers. When summed, they provide an estimate of the number of fish ascending and passing Lower Granite Dam each season. Window counts were made for each species using direct in-person visual monitoring and video monitoring during daytime hours [@Hatch1994]. Direct visual monitoring occurred during peak run times (April 1 – October 31) for 16 hours per day (0400 – 2000 hours). Video monitoring occurred during the beginning and tail ends of the adult runs (March 1 – March 31 and November 1 – December) for 10 hours per day (0600 – 1600 hours) [@USACE2015]. During direct visual monitoring, observers recorded each adult (≥ 30cm), by species, passing the window for 50 minutes of each hour of operation. The sum of the daily 50-minute counts were then multiplied by 1.2 to account for the 10 minutes when fish were not counted. Daytime window counts were not expanded for fish that may have ascended the ladder outside of operational hours (i.e., nighttime passage) [@USACE2015]. Window counts were accessed through the Columbia Basin Research Data Access in Real Time (DART) website, [www.cbr.washington.edu/dart/query/adult_daily](www.cbr.washington.edu/dart/query/adult_daily), using their window count query. Counts were provided for each day the fish ladder was open to passage, and had already been expanded by 1.2 (to account for the counting during 50 minutes of each hour).

@@

All captured fish were anesthetized, identified to species, examined for existing marks or tags, measured for fork length, and visually identified as wild or hatchery. The most widely used marking of hatchery fish is a clipped (removed) adipose fin, although coded wire tags are used in less than 10% of the hatchery releases. Some subset of hatchery fish are either intentionally or unintentionally released without a clipped adipose fin, and these are referred to as unclipped hatchery fish, or hatchery no-clips (HNC). For adults with intact adipose fins (unclipped), which includes wild and hatchery no-clip individuals, either a portion or all of the fish trapped (depending on the year) had scale and genetic tissue samples taken. Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012] and estimate the location of origin of wild fish using genetic stock identification [e.g., @Hargrove2019]. Prior to 2013, only fish determined to be wild in origin at the trap were sampled for scale and genetics. Starting in 2013, every unclipped sp/sum Chinook Salmon and steelhead trapped at Lower Granite Dam was genotyped to simplify collaborative logistics and better estimate the proportion of unclipped hatchery fish that appear phenotypically wild. @Camacho2018 provide further details on trap sample rates and valid sample selection. Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag. Final determination of wild, clipped hatchery, or unclipped hatchery origins were assigned using a post-hoc analysis of marks and tags, including parentage-based tagging based on the genetic samples taken [@Steele2013; @Steele2019]. Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game [@Camacho2018].

@@

The number of fish counted at the window, $Y^W_t$, is modeled as a (potentially) over-dispersed negative binomial process, with an expected value of $X_t^{day}$, the number of fish crossing the dam while the window is open. This is simply the total number of fish crossing that week, $X_t$, multiplied by the proportion of fish crossing while the window is open for counting, $\theta_t$. In the formula below, $p_t$ is the probability parameter and $r$ is the size parameter of a negative binomial distribution (as defined in @R-rjags). If $r$ is estimated to be small it provides evidence for over-dispersion, and as it grows very large, the negative binomial distribution behaves like a Poisson distribution.

@@

Our objective was to assess STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement for accuracy and precision, given a known simulated "truth", under different possible conditions. We developed those conditions from the combinations of two trap rate scenarios (constant and shut down for 3 weeks), two fallback/re-ascension and nighttime passage combinations (equal and unequal), and three window count error rates (none, low and high); resulting in twelve different scenarios (Table \@ref(tab:lgr-sim-tab)). The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at Lower Granite Dam between 2010-2015. We generated 99 simulations for each scenario, and ran STADEM on each one, resulting in a total of 1,188 model runs. Further details about the simulation procedures can be found in [Appendix A](#append1).

To assess the performance of the model, we evaluated the estimates of origin-specific escapement by several measures. Accuracy was captured by the relative bias: the difference between the simulated "truth" and the STADEM estimate, divided by the simulated "truth". We estimated the precision by examining the average coefficient of variation (CV) of the estimates. We calculated the root mean squared error (RMSE) as the square root of the mean of the squared bias in the estimate, which is equivalent to the square root of the variance of the estimator plus the square of the expected bias. Finally, we evaluated whether estimates of uncertainty were appropriate by calculating coverage probabilities by determining what proportion of the model results generated an estimated 95% credible interval that contained the simulated true value. To compare the STADEM results directly to window counts, we compared the model estimates of total unique fish and the simulated window counts to the simulated number of unique fish.

@@

STADEM results showed similar patterns for estimates of unique total fish and unique wild fish, in that both were unbiased with similar variance in the estimator across all scenarios as seen in the consistent interquartile distance in the boxplots of Figure \@ref(fig:rel-bias-lgd). For unique total fish, window counts were unbiased, except in scenarios when night passage and re-ascension rates were not equal (N-R). In those scenarios, they were biased by nearly 5% (Figure \@ref(fig:rel-bias-lgd)), the difference between night passage and re-ascension rates (Table \@ref(tab:lgr-sim-tab)). The variance of the window counts also grew as the simulated observer error grew (no error vs. Err_L vs. Err_H), whereas the variance in the STADEM estimates remained fairly constant, albeit often larger than the window count variance.

Estimates of wild escapement were unbiased, with an average relative bias of 0.2–0.3%, although the bias appeared equally distributed between positive and negative values across the simulations (Figure \@ref(fig:rel-bias-lgd)). The CV of the estimates averaged 2.0–3.0%, with higher CV's being associated with scenarios when the trap was closed for 3 weeks. The coverage probabilities always exceeded 95% across all scenarios. The RMSE was near 500 for each scenario, representing an estimate within 2% of the true value, demonstrating the accuracy and precision of STADEM (Table \@ref(tab:summ-tab)).

@@

Weekly nighttime passage and re-ascension proportions were significantly different (2-sample test of proportions, *p* < 0.05) in approximately 50% of the weeks with sufficient PIT tag data to test. The differences in the estimated rates were also apparent visually (Figure \@ref(fig:night-reasc-diff-fig)). In particular, there were several weeks when the window counts were quite large, but the nighttime passage and re-ascension rates differed by as much as 10%.

@@

The results from our simulations demonstrate that STADEM provides consistent, unbiased estimates of escapement across a range of scenarios. Regardless of whether observer error at the window is negligible, low or high, whether nighttime passage and re-ascension rates are equal or not, whether the trap is shut down for several weeks in a season or not, STADEM produced estimates of unique escapement that were unbiased, precise (CVs in the 2-3% range) and contained appropriate credible intervals. Although the average bias was often positive across our simulation scenarios (Table \@ref(tab:summ-tab)), Figure \@ref(fig:rel-bias-lgd) shows that some simulations resulted in negative biases, while some were positive. We posit that conducting further simulations would bring the average bias closer to zero, but run times constrained how many simulations we were able to conduct for each scenario.

As for application at Lower Granite Dam, the STADEM weekly estimates were always close to the window counts and the trap estimates, indicating that both sources of data were providing information about the same phenomenon. The STADEM weekly estimates were often slightly lower than those based on the number of fish caught in the adult trap because the trap estimates do not account for fallback and re-ascension (Figure \@ref(fig:jags-fit-fig)). However, the model's uncertainty is always smaller than the uncertainty from the trap estimates alone, whereas the window counts by themselves provide no estimate of uncertainty. When comparing the total estimates of each season with the window counts (Table \@ref(tab:lgr-tab)), STADEM estimates were higher when night passage rates exceeded re-ascension rates, and lower when the re-ascension rate was higher than the night passage rate. Given that there were several weeks with large window counts when the rates differed by as much as 10% (Figure \@ref(fig:night-reasc-diff-fig)), the counts for those weeks should be considered biased. ***For weeks when very few fish were caught in the trap or there was more uncertainty about the trap rate, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig).***

@@

STADEM is an improvement on window counts alone for several reasons. The first is that by accounting for night passage and re-ascension rates that are sometimes unequal, it corrects any bias in window counts. That bias ranged from -9.9% to 11% across all species and year combinations we fit STADEM to at Lower Granite Dam (Table \@ref(tab:lgr-tab)). Whether that amount of potential bias is of concern to management agencies is a decision for them, but we have presented a tool to correct that bias. The other justification for STADEM over window counts is that it provides an estimate of observer error for the window counts, leading to estimates of precision for unique fish which are unobtainable with window counts alone. Capturing and accounting for known sources of error is prudent to minimize management decision risk [@Harwood2003; @Ascough-Ii2008]. At the same time, STADEM's precision is still quite good, with CV's near 2-3%, well below the goal of 15% set forth by NOAA [@Crawford2011].

By incorporating data from the adult fish trap with live and video window counts in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly (for example, see Figure \@ref(fig:jags-fit-fig) when the trap was shut down for several weeks in July and August). At Lower Granite Dam, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently in five of the last ten years, often during peak run times [@USACE2010; @USACE2011; @USACE2012; @USACE2013; @Ogden2016a; @USACE2015; @USACE2016; @USACE2017; @USACE2018; @USACE2019]. Trap closures are typically associated with elevated water temperatures resulting in potential fish handling stress and/or trap malfunctions [@Ogden2016b]. Given predicted Pacific Northwest climate change scenarios [@Zhang2019] trap closures from high water temperatures may become more commonplace in the future, amplifying the need for a modeling framework that accounts for periods of missing data while still capturing uncertainty in the estimates. Additionally, having a framework in place that accounts for missing periods of data will allow for increased logistic flexibility if, for example, maintenance or construction is needed at the observational window or adult trap.

STADEM could be modified and run on a weekly basis or in near real-time to provide better in-season estimates for fisheries managers. Currently, the only roadblock to this at Lower Granite Dam is the identification of hatchery no-clip fish using genetic tissue samples [@Steele2013; @Steele2019] collected at the adult trap, which currently is completed post-hoc after the trapping season. The inclusion of genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement. However, if in-season management decisions do not require this correction or could accept the potential bias, origin calls at the trap could be used in-season as a first approximation to escapement. Final post-hoc estimates parsed by origin could then be completed at season’s end. All other data included in this model (e.g., window counts and PIT observations) are otherwise provided in near real-time by DART. Provided the Lower Granite Dam adult fish trap database was updated and available in near real-time, there are minimal obstacles for adapting the STADEM framework to provide in-season estimates of escapement.

Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate escapement of sp/sum Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@IPTDSW2020; @Kinzer2020]. Estimates of species and origin-specific escapement at Lower Granite Dam, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates [e.g., @Camacho2018; @Schrader2013] that are important for fisheries management and productivity monitoring of wild populations. As an example, STADEM estimates of unique wild fish at Lower Granite are combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @Waterhouse2020, to estimate escapement to Snake River populations and locations throughout the basin [@Orme2019]. Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations. With sex and age data collected at the adult fish trap [@Hargrove2019], this approach provides necessary information to evaluate productivity and population viability for select Snake River sp/sum Chinook Salmon and steelhead groups [@IPTDSW2020].

With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall-run Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided there exists a fish counting mechanism, a trap that can be used to sample a portion of the run, and tag detection infrastructure (e.g., a PIT tag detection array or similar). Justification and infrastructure exist for applying a modified STADEM framework for fish passing Bonneville Dam, the lowest dam on the Columbia River, or Priest Rapids Dam in the upper Columbia River. Both locations currently trap a sub-sample of passing Chinook Salmon and/or steelhead for biological information and use window counts as a surrogate of true escapement. However, each has at least some similar problems to those observed at Lower Granite Dam such as unaccounted observer and sampling error, nighttime passage, and/or re-ascension. Certainly, estimating an unbiased total return to the entire Columbia River basin (i.e., Bonneville Dam) and Upper Columbia River with uncertainty would benefit managers and decision making.

@@

```{r dag-fig, out.width = "100%", fig.cap = "Directed acyclic graph showing the STADEM model framework."}
include_graphics("../figures/ModelSchematic.png")
```
